Towards Compiler-Agnostic Performance in Finite-Difference Codes. (TBC)

### 2014-11-18 docs/papers/shallow_performance/shallow_perf.tex
Extending discussion of results.

Beginning with the vanilla \psykal version, we apply a series of
optimisations with the aim of recovering, as much as is possible, the
structure and thus performance of the original version of the code.

\begin{enumerate}

@@

\end{enumerate}

These optimisations do not always result in improved performance (AND
HENCE WE SHOULD CALL THEM TRANSFORMATIONS PERHAPS?). Whether or not
they do so depends both on the compiler and the problem size.

@@

In figure~\ref{FIG_orig_perf_summary} we plot the performance of the
original version of the shallow code for the range of compilers and
hardware considered here. This summary demonstrates the effect of the
larger last-level cache of the Intel Ivy Bridge system compared to
that of the Haswell system; note the significant drop in performance
in going from a problem size of $256^{2}$ to a problem size of
$512^{2}$ for the black and yellow bars. The performance of the Ivy
Bridge system (red and purple bars) only drops to this level when the
domain size is increased to $1024^{2}$. At this point, the working set
no longer fits within cache and performance is dominated by the
bandwidth to main memory, making it relatively insensitive to the
choice of compiler.

For problem sizes that fit within cache there is little to choose
between the performance obtained from the Intel and Gnu compilers on
the Intel Haswell CPU.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=2.8in]{best_psykal_summary}
\caption{Summary of the best performance achieved by any \psykal 
version of shallow with each of the compilers and CPUs under 
consideration.}
\label{FIG_best_psykal_perf_summary}
\end{figure}

@@

In Table~\ref{TABLE_opt_breakdown} we show detailed timings for the
$256^{2}$ case after each of the transformations has been applied to
the code. For the Gnu compiler, the vanilla \psykal version of shallow
is almost 20 times slower than the original. Copying the code for each
of the kernels into the module from which they are called has the most
dramatic effect on the performance of the compiled code. The next most
significant transformation is that of (manually) in-lining the kernel
code into the bodies of the loops from which they are called. This
gets the \psykal version to within 8\% of the performance of the
original.

\begin{table*}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{The time (s) taken to do 2000 steps of the $256^2$ case after each 
code transformation.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c|c|c|c}
\hline
Compiler:          &  Gnu    & Intel   & Gnu & Cray & Intel & IBM bgxlf \\
\hline
CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & PowerPC A2 \\
\hline
Original                             & 1.49 & 1.41 & 1.54 & 1.23 & 1.57  &        \\
Vanilla \psykal                      & 25.3 & 1.73 & 28.7 & 1.67 & 1.96  &        \\
Explicitly-sized arrays              & 25.3 & 1.72 & 28.6 & 1.39 & 1.93  & 113.74 \\
Kernels copied into module           & 2.20 & 1.73 & 1.84 & 1.42 & 1.93  &  68.28 \\
Loops fused                          & 2.18 & 1.58 & 1.79 & 1.47 & 1.76  &  71.91 \\
In-lined field-copy                  & 2.17 & 1.58 & 1.77 & 1.46 & 1.75  &  83.36 \\
Fused field-copy                     & 2.18 & 1.85 & 1.99 & 1.30 & 2.07  &  72.62 \\
In-line all kernels                  & 1.64 & 1.43 & 1.88 & 1.30 & 1.59  &        \\
Fully fuse 1st loop nest             & 1.62 & ADD  & 1.55 & 1.25 & 1.55  &        \\
Explicit, constant array/loop bounds & 1.61 & ADD  & 1.56 & 1.23 & 1.54  &        \\
\hline
\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 1.2 & 0.6 & 0.0 & -2.0 &     \\
\hline
\end{tabular}
\end{table*}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=85mm]{opt_stages_256}
\caption{The absolute performance of the code at each stage of 
optimisation for the $256^{2}$ domain.}
\label{FIG_opt_stages_256}
\end{figure}

@@

The authors would like to thank...

>>>>

Beginning with the vanilla \psykal version, we apply a series of
optimisations while obeying the \psykal separation of concerns, {\it
  i.e.} optimisation is restricted to the `middle layer' and leaves
the kernel and algorithm layers unchanged. The aim of these
optimisations is to recover, as much as is possible, the structure and
thus performance of the original version of the code. The
optimisations we have performed are as follows:
\begin{enumerate}

@@

\item Manually in-line kernel bodies into middle-layer code

\item Fully fuse first loop (only the outer loop of this loop nest was
  fused in the first pass)

\item Make the array/loop bounds explicitly constant for the
      duration of the time-stepping loop.
\end{enumerate}

We shall see that these optimisations do not always result in improved
performance. Whether or not they do so depends both on the compiler
and the problem size. We also emphasise that the aim of these
optimisations is to recover, as far as is possible, the structure of
the original version of the code. It may well be that transforming the
code into some other structure would result in better performance on a
particular architecture. However, exploring this optimisation space is
beyond the scope of the present work.

@@

In Figure~\ref{FIG_orig_perf_summary} we plot the performance of the
original version of the shallow code for the range of compilers and
hardware considered here. This summary demonstrates the effect of the
larger last-level cache of the Intel Ivy Bridge system compared to
that of the Haswell system; note the significant drop in performance
in going from a problem size of $256^{2}$ to a problem size of
$512^{2}$ for the black and green bars. The performance of the Ivy
Bridge system with the Cray or Intel compiler (red and purple bars)
only drops to this level when the domain size is increased to
$1024^{2}$. At this point, the working set no longer fits within cache
and performance is dominated by the bandwidth to main memory, making
it relatively insensitive to the choice of compiler.

@@

For this compiler-friendly form of the code, there is little
performance difference between the executables produced by the Gnu and
Intel compilers. However, the executable produced by the Cray compiler
generally performs significantly better. {\bf Why is this; can we work
out what it's doing that the others aren't?}

Note that the results for the Gnu compiler on Ivy Bridge show the same
drop-off (as problem size is increased) as those on the Haswell
system. This may be because Archer's login nodes seem to be running on
older hardware (E5-2650 with 20MB LLC) during the Archer upgrade. {\bf
  Need to re-run these cases once Archer is back up.}

Moving now the to the \psykal version of shallow,
Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
fastest \psykal version for each of the compiler/CPU combinations. The
only significant difference between this and the plot of the
performance of the original version of the code in
Figure~\ref{FIG_orig_perf_summary} is that the difference between the
Gnu- and Intel-compiled executables is slightly more marked. This is
clarified in Figure~\ref{FIG_slowdown_summary} which plots the
percentage difference between the performance of the original and the
(best) \psykal versions of shallow for each compiler/CPU combination.
It is clear from this plot that the difference in the performance of
the Gnu- and Intel-compiled excutables on the Haswell system is due to
the Gnu compiler being unable to recover the performance achieved by
the original code. In contrast there is no such problem for the
Gnu-compiled executable on the Ivy Bridge system. We attribute this to
the more recent version of the compiler on the latter system (4.9.1 as
opposed to 4.8.2).

\begin{figure}[!t]
\centering
\includegraphics[width=7.5cm]{best_psykal_summary}
\caption{Summary of the best performance achieved by any \psykal 
version of shallow with each of the compilers and CPUs under 
consideration.}
\label{FIG_best_psykal_perf_summary}
\end{figure}

@@

In fact, for the Gnu and Cray compilers on the Ivy Bridge system, the
\psykal version of shallow is never more than 2\% slower than the
original and in some cases is faster.  This demonstrates that we can
reliably recover the performance of the original version of the code,
despite the significant restructuring required by the \psykal
approach. {\bf We need to re-run/check the Intel results on Ivy Bridge
  once Archer is back from its upgrade.}

Having show that, in general, we can recover and occasionally improve
upon the performance of the original version of shallow, the next
logical step is to examine the necessary code transformations.  We do
this for the $256^{2}$ case since this fits within cache on both the
Haswell and Ivy Bridge CPUs.  Table~\ref{TABLE_opt_breakdown} shows
detailed timings for this case after each transformation has been
applied to the code. The same data is visualised in
Figure~\ref{FIG_opt_stages_256} where the performance of the best
\psykal version for a given compiler/CPU is given relative to the
performance of the original with the same compiler/CPU. For the Gnu
compiler, the vanilla \psykal version of shallow is almost 20 times
slower than the original. Copying the code for each of the kernels
into the module from which they are called has the most dramatic
effect on the performance of the compiled code. The next most
significant transformation is that of (manually) in-lining the kernel
code into the bodies of the loops from which they are called. This
gets the \psykal version to within 8\% of the performance of the
original.

\begin{table*}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{The time (s) taken to do 2000 steps of the $256^2$ case after each 
code transformation.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c|c|c|c}
\hline
Compiler:          &  Gnu    & Intel   & Gnu & Cray & Intel & IBM bgxlf \\
\hline
CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & PowerPC A2 \\
\hline
Original                             & 1.49 & 1.41 & 1.54 & 1.23 & 1.57  &        \\
Vanilla \psykal                      & 25.3 & 1.73 & 28.7 & 1.67 & 1.96  &        \\
Explicitly-sized arrays              & 25.3 & 1.72 & 28.6 & 1.39 & 1.93  & 113.74 \\
Kernels copied into module           & 2.20 & 1.73 & 1.84 & 1.42 & 1.93  &  68.28 \\
Loops fused                          & 2.18 & 1.58 & 1.79 & 1.47 & 1.76  &  71.91 \\
In-lined field-copy                  & 2.17 & 1.58 & 1.77 & 1.46 & 1.75  &  83.36 \\
Fused field-copy                     & 2.18 & 1.85 & 1.99 & 1.30 & 2.07  &  72.62 \\
In-line all kernels                  & 1.64 & 1.43 & 1.88 & 1.30 & 1.59  &        \\
Fully fuse 1st loop nest             & 1.62 & 1.44 & 1.55 & 1.25 & 1.55  &        \\
Explicit, constant array/loop bounds & 1.61 & 1.42 & 1.56 & 1.23 & 1.54  &        \\
\hline
\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 0.8 & 0.6 & 0.0 & -2.0 &     \\
\hline
\end{tabular}
\end{table*}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=90mm]{opt_stages_256}
\caption{The performance of the \psykal version of shallow for the
  $256^{2}$ domain at each stage of optimisation. Results are given as
  a percentage of the performance of the original code for each
  compiler/CPU combination. A figure greater than 100\% indicates that
  the \psykal version performs better than the original.}
\label{FIG_opt_stages_256}
\end{figure}

@@

This work made use of the ARCHER UK National Supercomputing Service
(\url{http://www.archer.ac.uk}).
